---
title: "Microbial Speciation with Extinction"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

# Setup Work Environment

```{r}
rm(list=ls())
getwd()
setwd("~/GitHub/MicroSpeciation/")

# Load packages
require("png")
require("grid")
````

# Load data
```{r}
ext <- read.table("~/GitHub/MicroSpeciation/data/Extinction_Rhode_Muller_2005.txt", sep = "\t", header = T)

ext[is.na(ext)] <- 0 # assign missing values a background extinction rate
```

# Discrete logistic growth
```{r}
# Speciation parameters
r <- 0.0067 # speciation rate
S <- 1 # initial number of species
K <- Inf # carrying capacity

# Extinction parameters
Eb <-rnorm(4000, mean = 1, sd = 2) # baseline extinction (%)
Eb <- ifelse(Eb<0,0.1,Eb)
Ee <- ext$Ext[1:4000] # mass extinction events sensu Sepkoski
GOE <- rnorm(2426-2240, mean = 50, sd = 10) # great oxidation event
GOE <- ifelse(GOE<0,0,GOE)
GOE <- ifelse(GOE>90, 90, GOE)
GOE <- c(Ee[1:(4000-2426)],GOE,Ee[1761:4000])
Ee <- GOE
Eb.t <- 0.1 # initial Eb
Ee.t <- 0.1 # initial Ee

# Richness with discrete logistic with extinction
for(time in 2:4000) {
  S[time] = S[time-1] + (S[time-1]*r) * (1-S[time-1]/K) # growth
  - (S[time-1] * Eb) # baseline extinction
  - (S[time-1] * Ee) # extinction events
}

# Number of taxa lost from baseline extinction
for(time in 2:4000) {
  Eb.t[time] = S[time-1] * Eb[time]
}

# Number of taxa lost from mass extinction events
for(time in 2:4000) {
  Ee.t[time] = S[time-1] * Ee[time]
}

# Dataframe with times, richness, and extinctioins
E.total <- Eb.t + Ee.t
rich <- data.frame(ext$time[1:4000], ext$MYA[1:4000], Eb, Ee, Ee+Eb, Eb.t, Ee.t, E.total, S)
colnames(rich) <- c("time", "mya", "E.per.b", "E.per.e", "E.per.tot", "E.b.rich", "E.e.rich", "E.total.rich", "S")
```

```{r}
png(filename="~/GitHub/MicroSpeciation/figures/logistic.richness.png",
        width = 1200, height = 800, res = 96*2)

layout(matrix(c(1:2), byrow = T))
par(mar = c(0.5, 4, 1, 1), oma = c(3, 4, 1, 1))

# Extinction panel
plot(rich$mya, rich$E.per.b, type = "l", log = "x", col = "red", ylim = c(0, 75), xlim = c(5000, 1), yaxt = "n", lty = 2, xaxt = "n", las = 1, ylab = "", xlab = "", lwd = 2)

points(rich$mya, rich$E.per.e, type = "l", col = "blue", lwd = 2)

box(lwd = 2)

axis(side = 2, lwd.ticks = 2, cex.axis = 1.25, las = 1,
    labels = T, at = c(0, 25, 50, 75))

axis(side = 4, lwd.ticks = 2, cex.axis = 1.25, las = 1,
   labels = F, at = c(0, 25, 50, 75))

axis(side = 1, lwd.ticks = 2, cex.axis = 1.25 , las = 1,
    labels = F, at = c(4000, 400, 40, 1))

axis(side = 3, lwd.ticks = 2, cex.axis = 1.25, las = 1,
    at = c(4000, 400, 40, 1), labels = F)

mtext(side = 2, "Extinction (%)", line = 3, cex = 1.5)
#mtext(side = 1, "Time (mya)", line = 3.5, cex = 1.5)

legend(15, 65, c("mass", "baseline"), lty = c(2,2), col = c("blue", "red"),
       lwd = 2.5, seg.len = 2, bty = 'n', pt.cex = 2)

# Richness panel
plot(rich$mya, rich$E.total.rich, type = "l", log = "yx", col = "red", ylim = c(1, 10^13), xlim = c(5000, 1), yaxt = "n", lty = 2, xaxt = "n", las = 1, ylab = "", xlab = "", lwd = 2)

points(rich$mya, rich$S, type = "l", col = "blue", lwd = 2)

box(lwd = 2)

axis(side = 2, lwd.ticks = 2, las = 1, cex.axis = 1.25, 
   labels = c("0", "4", "8", "12"), 
   at = c(10^0, 10^4, 10^8, 10^12))

axis(side = 4, lwd.ticks = 2, cex.axis = 1.25, las = 1,
   labels = F, at = c((10^0, 10^4, 10^8, 10^12))

axis(side = 1, lwd.ticks = 2, cex.axis = 1.25 , las = 1,
    labels = T, at = c(4000, 400, 40, 1))

axis(side = 3, lwd.ticks = 2, cex.axis = 1.25, las = 1,
    at = c(4000, 400, 40, 1), labels = F)

mtext(side = 1, "Time (mya)", line = 2.5, cex = 1.5)
mtext(side = 2, "log10(Richness)", line = 3, cex = 1.5)

legend(15, 10^6, c("extant", "extinct"), lty = c(2,2), col = c("red", "blue"),
       lwd = 2.5, seg.len = 2, bty = 'n', pt.cex = 2)


# Close Plot Device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("~/GitHub/MicroSpeciation/figures/logistic.richness.png")
grid.raster(img)
````


# Mass extinction
  # Great Oxidaton Event (http://www.pnas.org/content/114/8/1811)
rect(2426, min(rich$S), 2240, max(rich$S), 
     border = "gray", col = "gray")
  # Ordovician–Silurian transition
rect(450, min(rich$S), 440, max(rich$S), 
     border = "gray", col = "gray")
  # Late Devonian extinction
rect(375, min(rich$S), 360, max(rich$S), 
     border = "gray", col = "gray")
  # Permian–Triassic transition
rect(252, min(rich$S), 252, max(rich$S), 
     border = "gray", col = "gray")
  # Triassic-Jurassic extinction
rect(201.3, min(rich$S), 201.3, max(rich$S), 
     border = "gray", col = "gray")
  # Cretaceous-Paleogene
rect(66, min(rich$S), 66, max(rich$S), 
     border = "gray", col = "gray")




# Richness panel
plot(rich$mya, rich$S, type = "l", log = "xy", col = "red", xlim = c(max(rich$mya), min(rich$mya)), xlab = "Time (mya)", 
     las = 1, ylab = "Species Richness")

# Mass extinction
  # Great Oxidaton Event (http://www.pnas.org/content/114/8/1811)
rect(2426, min(rich$S), 2240, max(rich$S), 
     border = "gray", col = "gray")
  # Ordovician–Silurian transition
rect(450, min(rich$S), 440, max(rich$S), 
     border = "gray", col = "gray")
  # Late Devonian extinction
rect(375, min(rich$S), 360, max(rich$S), 
     border = "gray", col = "gray")
  # Permian–Triassic transition
rect(252, min(rich$S), 252, max(rich$S), 
     border = "gray", col = "gray")
  # Triassic-Jurassic extinction
rect(201.3, min(rich$S), 201.3, max(rich$S), 
     border = "gray", col = "gray")
  # Cretaceous-Paleogene
rect(66, min(rich$S), 66, max(rich$S), 
     border = "gray", col = "gray")

# Close Plot Device
dev.off()
graphics.off()

# Show Plot
img <- readPNG("~/GitHub/MicroSpeciation/figures/logistic.richness.png")
grid.raster(img)
```

# Discrete logistic growth
```{r}
log.growth <- function(R, S, K) {
	Snext <- R * S * (1-S/K) + S
	return(Snext)
}

# log.growth.ext <- function(R, S, K) {
# 	Snext <- (R*S*(1-S/K) + S) * E
# 	return(Snext)
# }

R <- 0.0067 # intrinsic growth rate
S <- ext$time # range of S corresponding to times
K <- 50 # carrying capacity
E <- ext$Ext # extinction events

St <- log.growth(R, S, K) #calculate S + 1 across ranage of S

# St.ext <- log.growth.ext(R, S, K, E) #calculate N+1 across ranage of N for Logistic

plot(ext$MYA, St, type = "l", col = "red", xlim = rev(range(ext$MYA)), xlab = "Time (mya)", 
     las = 1, ylab = "Species Richness")



N0 <- 1
K <-10^12
rmax <- 0.0067
C <- K/N0-1 
#first generate a list of time indices
t <- ext$time 
N <- K/(1+C*exp(-rmax*t))
#make it pretty
data.frame(t=t,N=N)
plot(t,N,type="l")

http://cnr.lwlss.net/DiscreteStochasticLogistic/
simDSLogistic=function(K,r,N0){
  # Unusually, for this model, we know the number of events a priori
  eventNo=K-N0
  # So we can just generate all required random numbers (quickly) in one go
  unifs=runif(eventNo)
  # Every event produces one cell and consumes one unit of nutrients
  clist=(N0+1):K
  # Simulate time between events by generating 
  # exponential random numbers using the inversion method
  dts=-log(1-unifs)/(r*clist*(1-clist/K))
  return(data.frame(t=c(0,cumsum(dts)),c=c(N0,clist)))
}



r <- 1.0067
S <- 1
K <- 10
for(time in 2:4000) {
  S[time]=S[time-1]*r
}

r <- 0.67
S <- 1
K <- 5

for(time in 2:40) {
  S[time] = S[time-1] + (S[time-1]*r) * (1-S[time-1]/K)
}

r <- 0.67
S <- 1
K <- 1000
Eb <- 0.1
Ee <- ext$Ext[2:40]

for(time in 2:40) {
  S[time] = S[time-1] + (S[time-1]*r) * (1-S[time-1]/K) # growth
  - ((S[time-1]*r) * (1-S[time-1]/K) * Eb) # baseline extinction
  - ((S[time-1]*r) * (1-S[time-1]/K) * Ee) # extinction events
}


```

